---
title: 'DS202: Lab 4'
author: "Muhammad Khairi Norizan"
date: "10/26/2021"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)

defense <- read_excel('./cyclonesFootball2020.xlsx', sheet='Defensive')
offense <- read_excel('./cyclonesFootball2020.xlsx', sheet='Offensive')
biography <- read_excel('./cyclonesFootball2020.xlsx', sheet='Biography')

defClean <- data.frame(defense)
offClean <- data.frame(offense)
bioClean <- data.frame(biography)
```

## Part 1: Cleaning Data

1. The player names and opponent names should be factors.
```{r}
defClean$Name <- as.factor(defClean$Name)
offClean$Name <- as.factor(offClean$Name)
```

2. All offensive and defensive statistics should be numerical. The `Weight` column in biography should also be numeric. (Optional: Read `?across` and learn to mutate multiple columns simultaneously using the `across` function in `dplyr`)
```{r}
defClean <- defClean %>% mutate(across(Tackles_Solo:Pass_PB, as.numeric))

offClean <- offClean 

offClean <- offClean %>% 
  separate(Passing_CMPATT, c('Passing_CMP', 'Passing_ATT'), sep = "-")
offClean <- offClean %>% mutate(across(Receiving_REC:Passing_INT, as.numeric))
bioClean$Weight <- as.numeric(biography$Weight)
```

3. The `Height` column in `biography` should be numeric. (Hint: Use a unit that works well)
```{r}
bioClean <- bioClean %>% 
            separate(Height, c('feet', 'inches'), "-", convert = TRUE) %>% 
            mutate(height = (12*feet + inches)*2.54) %>% select(-c('feet', 'inches')) %>% 
            rename('height(cm)'=height)
```

4. Add a column to both `defense` and `offense` indicating whether it is the first or the second game versus an opponent. We had a single game versus each opponent except for Oklahoma, which we played against twice. (Hint: You can use `group_by()`, `mutate()`, and `row_number()`)
```{r}
offClean <- offClean %>% 
  group_by(Name) %>% 
  mutate(FirstTime = !duplicated(Opponent_Opponent))

defClean <- defClean %>% 
  group_by(Name) %>% 
  mutate(FirstTime = !duplicated(Opponent_Opponent))
```

```{r}
print(str(offClean))
print(str(defClean))
print(str(bioClean))
```

## Part 2: Tidying and exploring
1. Reformat the `offClean` data frame into a tidy format using pivot_longer. The type of statistic (Receiving_REC for the number of receiving, Receiving_YDS for receiving yards, etc) is added as a new key column named `stat`.
```{r}
offClean <- offClean %>% pivot_longer(Receiving_REC:Passing_INT, names_to = "stat", values_to = "Value")
print(offClean)
```

2. Obtain the player-level offensive statistics. Namely, for each player and each type of statistics, obtain the sum of the statistics in all games
```{r, warning=FALSE}
playerOffenseStat <- offClean %>% 
                      group_by(Name, stat) %>% 
                      summarize(statSum = sum(Value, na.rm = TRUE))

print(playerOffenseStat)
```

3. Show the distributions of the player-level offensive statistics. The distribution of each offensive statistic should be shown in a separate facet. Explain the patterns you see.
```{r, warning=FALSE}
ggplot(playerOffenseStat, aes(x=statSum)) + 
  geom_histogram() + facet_wrap(~stat, scales = 'free')
```

From the histogram presented in the above, we can see that ISU football team is doing offensively better especially in receiving rec and receiving yds. These two facets (Receiving_REC and Receiving_YDS) tells that the are a few numbers of players has pretty far receiving yards and rec.

4. Did ISU have better offense against Oregon or Oklahoma? Answer this question by creating a scatterplot with the x- and y-axes being the receiving yards of the ISU players against each of these two opponents. A large number of receiving yards indicates good offense.
```{r, warning=FALSE}
oregonOklahoma <- offClean %>% 
                  filter(Opponent_Opponent == 'Oklahoma' | Opponent_Opponent == 'Oregon') %>% 
                  filter(stat == 'Receiving_YDS') %>% pivot_wider(names_from = Opponent_Opponent, values_from = Value)

ggplot(oregonOklahoma, aes(Oklahoma, Oregon)) + geom_point() + coord_fixed(xlim = c(0,100), ylim = c(0,100)) + geom_abline(intercept = 0, slope = 1)
```

The graph above separates the receiving yards of the ISU players data against Oregon and Oklahoma. The data below the line shows the performance of ISU players against Oklahoma, and the data above the the performance against Oregon. Since there are more data below the line, therefore, we can say that ISU players perform better against Oklahoma in terms of receiving yards.

5. Separate the `Hometown` column into two columns, namely, the city and the state where the player comes from. Print the first few rows of these two columns. (Hint: look at the `sep=` argument of `separate`)
```{r}
bioClean <- bioClean %>% 
            separate(Hometown, c('city', 'state'), sep = ",", convert = TRUE)

cityStateBio <- bioClean %>% select(city, state)
head(cityStateBio)
``` 
 
6. How many players are there from each state?
```{r}
playersStateCount <- bioClean %>% 
                      group_by(state) %>% 
                      summarize(playersCount = n())

print(playersStateCount)
```

7. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team? Comment on both the defense and offense.
```{r, warning=FALSE}
offClean %>% 
  mutate(PurdyStat = ifelse(Name == 'Purdy, Brock', Value, NA)) %>% 
  ggplot() +
  geom_boxplot(aes(x = stat, y = Value)) +
  geom_point(aes(x = stat, y = PurdyStat), color = 'Red') + 
  coord_flip() + facet_wrap(~Opponent_Opponent)
```

On the offensive point of view, Brock Purdy has done better than the average ISU team. Looking at the graph above, Brock Purdy has done an excellent job in passing yards for all the opponent he against with. 

```{r, warning=FALSE}
# Transform defClean to a tidy data set
defClean <- defClean %>% pivot_longer(Tackles_Solo:Pass_PB, names_to = "stat", values_to = "Value")

defClean %>% 
  mutate(PurdyStat = ifelse(Name == 'Purdy, Brock', Value, NA)) %>% 
  ggplot() +
  geom_boxplot(aes(x = stat, y = Value)) +
  geom_point(aes(x = stat, y = PurdyStat), color = 'Red') + 
  coord_flip() + facet_wrap(~Opponent_Opponent)
```

On the other hand, Brock Purdy is defensively poor compared to the other team. However, Brock Purdy has done slightly better in solo tackle against Texas Tech compared when he is playing against a different team. 